---
title: "Alabama Covid-19 Cases"
author: "Tom Plunkett"
date: "2025-03-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:

I live in Huntsville, Alabama so I used R to analyze Covid-19 cases for the State of Alabama.  I compared COVID cases per thousand with respect to deaths per thousand.  I used COVID-19 data sourced from the Johns Hopkins GitHub repository. These datasets provide daily updates on the total number of confirmed COVID-19 cases and deaths for each country worldwide, as well as for each state in the United States. The data sets can be found at the following URL: https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series.

## Libraries

```{r getData, message=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(readr)
```


## Importing Dataset and reading data from csv file

First, import the required dataset from the public repository from the following source: https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/


```{r}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read.csv(urls[3])
US_deaths <- read.csv(urls[4])

```

## TIDY AND TRANSFORM

Now tidy the dataset and transform it into useful form for further data analysis.


```{r}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>%
  select(-c(Lat,Long))

global <- global_cases %>% full_join(global_deaths) %>% 
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>%
  unite("Combined_Key", c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

global

summary(global)

```


### Removing character "X" and converting date to date data type

Now convert the date to a date data type in the US data set, removing the "X" character from the date. Also, we can rename the column.

```{r tidy_US_data, message=FALSE}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), 
                                      names_to = "date", 
                                      values_to = "cases") %>% 
  select(Admin2:cases) %>% 
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), 
                                      names_to = "date", 
                                      values_to = "deaths") %>% 
  select(Admin2:deaths) %>% 
  select(-c(Lat, Long_))

US_cases$date <- gsub("X", "", as.character(US_cases$date))
US_deaths$date <- gsub("X", "", as.character(US_deaths$date))

US <- US_cases %>% full_join(US_deaths) %>% 
  mutate(date = mdy(date))
  
```

### Organizing  the data set

Now organize our dataset as per our analysis by grouping all the countries in a new dataframe.


```{r country}
global_country <- global %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

global_country

```

```{r countrytotals}
total_all <- global_country %>% 
  group_by(Country_Region) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases / population, deaths_per_thou = 1000*deaths / population) %>% filter(cases>0, population>0)

total_all %>% slice_min(deaths_per_thou, n = 10)
```



### Visualization of COVID cases in Alabama

Data visualization is the graphical representation of information and data. First, visualize the COVID cases in Alabama using the absolute numbers of cases.

```{r more_US_data, message=FALSE}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()
```


Let us visualize the COVID cases that occurred in Alabama in the US from the US dataset. The below graphs show the number of cases and deaths in Alabama.  First let's visualize the absolute numbers.

```{r}
state <- "Alabama"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("Covid Cases in US state ", state), y = NULL)
```

Now lets visualize using log on the y axis.

```{r}
state <- "Alabama"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("Covid Cases in US state ", state), y = NULL)
```
## ANALYSING AND MODELING 

Let us now delve deeper into the data set by comparing COVID cases per thousand with respect to deaths per thousand. Here is a linear model to show the correlation between them.
The below  model roughly follows the pandemic trend in the region. The yellow curve shows the observed deaths per thousand cases. The black one represents the linear model’s estimate of cases per thousand.


```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = total_all)
summary(mod)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
total_pred <- total_all %>% mutate(pred = predict(mod))
```

```{r}
total_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "black") + 
  geom_point(aes(x = cases_per_thou, y = pred), color = "yellow")
```




## BIAS IDENTIFCATION:

Aggregating data from various sources introduces variability in quality, accuracy, and timeliness. The reliability of each source impacts the overall dataset. Inconsistent reporting of COVID cases and deaths across regions can skew the data. Varying definitions of COVID-related deaths adds complexity. Confirmed cases may not capture all infections due to testing limitations.

Researchers and data collectors may unintentionally introduce bias based on their perspectives.  Undiagnosed or asymptomatic cases may be unreported. This introduces uncertainty about the actual prevalence of COVID-19.

## CONCLUSION

The data visualizions showed how cases have increased over time.  Yet the success of the vacinnes have signficantly reduced the deaths that were present during the early phase of the pandemic.  

This John Hopkings Covid-19 data set helps researchers look for trends in the data on cases and death.  More insights can be found by digging deeper into this data set.  



```{r session}

sessionInfo()

```
